---
title: "Clustering"
author: "HN"
date: "2025-02-10"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
    df_print: paged
    theme: readable
    code_download: true
---
# This R Markdown document illustrates the analysis pipeline used to identify
# subphenotypes of sepsis-associated acute kidney injury (SAKI) using
# proteomic biomarker data.
#
# The pipeline includes:
# - ensemble clustering
# - cluster validation
# - survival analysis
# - pathway enrichment
# - figure generation


```{r setup, include=FALSE}
rm(list = ls()); gc()
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# --- Core tidy tools ---
library(tidyverse)

# --- Olink & Enrichment ---
library(OlinkAnalyze)
library(clusterProfiler)
library(enrichplot)
library(GO.db)
library(org.Hs.eg.db)
library(AnnotationDbi)

# --- Clustering & Validation ---
library(diceR)
library(tidyLPA)
library(cluster)
library(factoextra)
library(fpc)
library(clValid)
library(umap)

# --- Survival Analysis ---
library(survival)
library(survminer)

# --- Tables & Reporting ---
library(gtsummary)
library(gt)

# --- Plotting & Layout ---
library(pheatmap)
library(ggplot2)
library(ggplotify)
library(ggrepel)
library(ggnewscale)
library(ggforce)
library(ggfortify)
library(patchwork)
library(cowplot)
library(EnhancedVolcano)

# --- I/O ---
library(readxl)

# ------------------------------------------------------------
# Paths: this Rmd lives in code/, project root is one level up
# ------------------------------------------------------------
set.seed(42)
knitr::opts_knit$set(root.dir = normalizePath("..", winslash = "/"))
message("Working directory set to: ", getwd())

# Create results/ for all outputs
dir.create("results", showWarnings = FALSE)

# Helper to save ggplots into results/
fig_path  <- function(name) file.path("results", name)
save_plot <- function(p, name, w = 7, h = 5, dpi = 300) {
  ggplot2::ggsave(filename = fig_path(name), plot = p, width = w, height = h, dpi = dpi)
}


```

# 1) Read the data

```{r read-data}
outcomes <- c("death.28d","timetodeath.28d","death.90d","timetodeath.90d",
          "MAKE.7d","MAKE.90d","ICU_LoS","ICU_freedays.28d","ICU_freedays.90d","severeAKI.7d")


# Olink + clinical outcomes
adre_olink_MT <- read.csv("data/ADRENOSS_Olink_Clinical_Final_deid.csv")

# Quick NA scan (console)
suppressWarnings(print(colSums(is.na(adre_olink_MT[outcomes]))))

# Setup wide biomarker matrix (remove IDs/outcomes)
wide <- adre_olink_MT %>%
  dplyr::select(-c("Patient", all_of(outcomes)))

# biomarker list
biomarker_list_sig <- colnames(wide[,-1]) # keep your naming

# quick dims
suppressWarnings(print(dim(adre_olink_MT))); 
suppressWarnings(print(length(biomarker_list_sig))); 
suppressWarnings(print(dim(wide)))

```

# 2) Clustering

```{r running with k from 2 to 6 for choosing best k}
nk <- 2:6
reps <- 50
alg <- c("sc", "km", "pam", "ap", "gmm", "block", "cmeans", "hdbscan", "nmf", "hc", "som", "diana")

con_file <- "data/clustering/adre_ensemble_consensus_cluster_2to6.RDS"
if (file.exists(con_file)) {
  con_clust <- readRDS(con_file)
} else {
  con_clust <- consensus_cluster(
    data = as.matrix(wide[, -1]),
    nk = nk,
    reps = reps,
    algorithms = alg
  )
  saveRDS(con_clust , con_file)
}

eval_file <- "data/clustering//adre_ensemble_eval_wo_trim_2to6.RDS"
if (file.exists(eval_file)) {
  con_eval_rslt_wo_trim <- readRDS(eval_file)
} else {
  set.seed(42)
  con_eval_rslt_wo_trim <- consensus_evaluate(
    data = as.matrix(wide[, -1]), 
    con_clust, 
    k.method = NULL,
    plot = FALSE
  )
  saveRDS(con_eval_rslt_wo_trim , eval_file)
}

# PAC across k
adre_pac <- tryCatch(rowMeans(con_eval_rslt_wo_trim$pac[,-1]), error = function(e) NA_real_)
adre_pac
```

## 2.1 Consensus clustering (k = 2)

```{r clustering-k2}
nk <- 2
reps <- 50
alg <- c("sc","km","pam","ap","gmm","block","cmeans","hdbscan","nmf","hc","som","diana")

con_file <- "data/clustering/adre_ensemble_consensus_cluster.RDS"
if (file.exists(con_file)) {
  con_clust <- readRDS(con_file)
} else {
  con_clust <- consensus_cluster(
    data = as.matrix(wide[, -1]),
    nk = nk,
    reps = reps,
    algorithms = alg
  )
  saveRDS(con_clust, con_file)
}
```

## 2.2 Evaluate (no trim) + keep algorithms

```{r evaluate-no-trim}
eval_file <- "data/clustering/adre_ensemble_eval_wo_trim.RDS"
if (file.exists(eval_file)) {
  con_eval_rslt_wo_trim <- readRDS(eval_file)
} else {
  set.seed(42)
  con_eval_rslt_wo_trim <- consensus_evaluate(
    data = as.matrix(wide[, -1]),
    con_clust,
    k.method = 2,
    plot = FALSE
  )
  saveRDS(con_eval_rslt_wo_trim, eval_file)
}

algorithms_kept <- con_eval_rslt_wo_trim$trim.obj$alg.keep
if (is.null(algorithms_kept)) algorithms_kept <- dimnames(con_clust)[[3]]


```

## 2.3 Cluster assignments

```{r assignments}
rds_file <- "data/clustering/adre_ensemble_assignments_deid.RDS"
if (file.exists(rds_file)) {
  cluster_results <- readRDS(rds_file)
} else {
  cluster_results <- list()
  ks <- 2:2
  ensemble_methods <- c("CSPA", "LCA")
  trim_levels <- "All"
  combinations <- tidyr::expand_grid(fn = ensemble_methods, trim = trim_levels)

  for (k in ks) {
    cluster_results[[as.character(k)]] <- apply(combinations, 1, function(x) {
      fn <- x[["fn"]]
      args <- list(E = con_clust[, , algorithms_kept, , drop = FALSE])
      if (fn == "CSPA") {
        args$k <- k
      } else {
        k_names <- dimnames(con_clust)[[4]]
        k_pos <- which(k_names == as.character(k))
        if (length(k_pos) == 0) return(NULL)
        args$E <- args$E[, , , k_pos, drop = FALSE]
      }
      cl_assign <- try(do.call(fn, args))
      if (inherits(cl_assign, "try-error")) return(NULL)
      setNames(data.frame(cl_assign), paste0(fn, ".", x[["trim"]]))
    })
  }

  cluster_results <- lapply(cluster_results, dplyr::bind_cols)
  cluster_results <- lapply(cluster_results, function(x) dplyr::bind_cols(wide["SampleID"], x))
  cluster_results <- cluster_results$`2` %>% data.frame()
  saveRDS(cluster_results, rds_file)
}

suppressWarnings(print(table(cluster_results$LCA.All)))
suppressWarnings(print(prop.table(table(cluster_results$LCA.All))))
```

## 2.4 Heatmap

```{r heatmap}
heatmap_plot <- wide %>% 
  column_to_rownames("SampleID") %>% 
  cbind(., Cluster = cluster_results$LCA.All) %>% 
  arrange(Cluster) %>% 
  dplyr::select(-any_of(c("Cluster"))) %>% 
  pheatmap::pheatmap(., 
           cluster_rows = FALSE, 
           gaps_row = sum(cluster_results$LCA.All == 1), 
           show_rownames = FALSE, 
           show_colnames = FALSE,
           color = colorRampPalette(c("blue", "white", "red"))(100))


# Convert to ggplot safely for patchwork later
if (!is.null(heatmap_plot$gtable)) {
  heatmap_plot_gp <- ggplotify::as.ggplot(heatmap_plot$gtable)
} else {
  # Rebuild gtable if needed
  heatmap_plot2 <- pheatmap::pheatmap(wide %>% column_to_rownames("SampleID"),
                                      cluster_rows = FALSE, show_rownames = FALSE, show_colnames = FALSE)
  heatmap_plot_gp <- ggplotify::as.ggplot(heatmap_plot2$gtable)
}

# Save
save_plot(heatmap_plot_gp, "heatmap.png", w = 7, h = 7)
```

## 2.5 Clustering indices + data prep for plots
```{r Clustering plot}
biomarker_list_sig <- colnames(wide[,-1])

adre_olink_MT_clustered <- adre_olink_MT %>% 
  inner_join(cluster_results, by = "SampleID") %>% 
  dplyr::select(c("SampleID", "CSPA.All", "LCA.All", all_of(biomarker_list_sig), all_of(outcomes)))

wide_clustered <- adre_olink_MT_clustered  %>% 
  dplyr::select(-c("SampleID", "LCA.All", "CSPA.All", all_of(outcomes))) %>%
  dplyr::select(where(is.numeric))
```

# 4) Analysis


## 4.1) COMBINED PLOT: PCA + Survival + MAKE90

### 4.1.1) Survival

```{r Survival plot adre}
cox_vars <- c("death.28d", "timetodeath.28d", "death.90d", "timetodeath.90d")



adre_cox_df <- adre_olink_MT %>% 
  dplyr::select(Patient, all_of(cox_vars)) %>% 
  inner_join(adre_olink_MT %>% dplyr::select(c(SampleID, Patient)), by = "Patient") %>% 
  inner_join(adre_olink_MT_clustered%>% dplyr::select(c(SampleID, LCA.All)), by = "SampleID") %>% 
  mutate(Cluster = factor(LCA.All, levels = c(1,2)),
         timetodeath.90d = case_when(
           death.90d == 0 & timetodeath.90d <= 90 ~ 90.001,
           TRUE ~ timetodeath.90d
         )) %>% 
  dplyr::select(-c(Patient, SampleID, LCA.All)) 

my_palette <- c("#002395", "#EF4135") 

adre_olink_MT_clustered <- adre_olink_MT_clustered %>% 
  mutate(Cluster = factor(LCA.All, levels = c(1, 2)))

fit_cl <- survfit(Surv(timetodeath.90d, death.90d) ~ Cluster, data = adre_cox_df)
cox_fit <- coxph(Surv(timetodeath.90d, death.90d) ~ Cluster,  data = adre_cox_df)

cox_summary <- summary(cox_fit)
hr <- round(cox_summary$coefficients[1, "exp(coef)"], 2)
ci_lower <- round(cox_summary$conf.int[1, "lower .95"], 2)
ci_upper <- round(cox_summary$conf.int[1, "upper .95"], 2)
p_val <- signif(cox_summary$coefficients[1, "Pr(>|z|)"], 2)
annotation_text <- paste0(
  "Log-rank p-value: ", ifelse(p_val < 0.0001, "< 0.001", format.pval(p_val, eps = 1e-3)),
  "\nHR = ", round(hr, 2), " [", round(ci_lower, 2), ", ", round(ci_upper, 2), "]"
)

sur_plot <- ggsurvplot(
  fit_cl,
  data = adre_cox_df,
  palette = my_palette,
  conf.int = TRUE,
  conf.int.alpha = 0.2,
  conf.int.style = "ribbon",
  censor = TRUE,
  censor.shape = 124,
  censor.size = 2,
  risk.table = TRUE,
  risk.table.col = "strata",
  risk.table.y.text = FALSE,
  risk.table.height = 0.3,
  break.time.by = 15,
  xlim = c(0, 90),
  pval = FALSE,
  size = 1,
  legend = "none",
  xlab = "",
  ylab = "Survival probability",
  ggtheme = theme_bw(base_size = 12),
  tables.theme = theme(
    text = element_text(size = 16, face = "bold", color = "black"),
    axis.text.x = element_text(size = 18, face = "bold", color = "black"),
    axis.text.y = element_text(size = 18, face = "bold", color = "black")
  ),
  risk.table.fontsize = 8
)

sur_plot$table <- sur_plot$table +
  labs(title = "Number at risk", y = NULL) +  
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(size = 16, face = "bold"),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none"
  )

sur_plot$plot <- sur_plot$plot +
  scale_y_continuous(labels = scales::percent_format()) +
  annotate("text", x = 5, y = 0.15, label = annotation_text, hjust = 0, size = 7, fontface = "bold", color = "black") +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6),
    axis.ticks = element_line(color = "black"),
    axis.text.x = element_text(color = "black", size = 14),
    axis.text.y = element_text(color = "black", size = 14),
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold"),
    plot.background = element_rect(fill = "white"),
    legend.position = "none"
  ) +
  labs(x = "Time (days)", y = "Survival probability")

sur_plot$table <- sur_plot$table +
  labs(title = "Number at risk", y = NULL) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(size = 16, face = "bold"),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    legend.position = "none"
  )

sur_plot
```

### 4.1.2) PCA

```{r PCA plot}
pca_result <- prcomp(adre_olink_MT_clustered %>% 
                       dplyr::select(-c(SampleID, CSPA.All, LCA.All, Cluster, all_of(outcomes))), scale = TRUE)

# Align PC1 with prior orientation
pca_result$x[, 1] <- -pca_result$x[, 1]
pca_result$rotation[, 1] <- -pca_result$rotation[, 1] 

pca_plot <- autoplot(
  pca_result,
  data = adre_olink_MT_clustered %>% mutate(Cluster = factor(Cluster)),
  colour = 'Cluster',
  shape = 'Cluster',
  frame = TRUE,
  frame.type = 'convex',
  frame.alpha = 0.2,
  size = 1.5
) +
  scale_color_manual(values = my_palette) +        
  scale_fill_manual(values = my_palette) +         
  scale_shape_manual(values = c(16, 17)) +        
  theme_classic() +
  theme(
    text = element_text(size = 14),
    legend.position = "none"
  )

pca_plot
```

### 4.1.3) MAKE90

```{r Donut plot for MAKE90}
donut_data <- adre_olink_MT_clustered %>% 
  dplyr::filter(!is.na(MAKE.90d)) %>% 
  mutate(MAKE.90d = ifelse(MAKE.90d == 1, "Y", "N")) %>% 
  count(Cluster, MAKE.90d) %>%
  group_by(Cluster) %>%
  mutate(Percentage = n / sum(n) * 100) %>%
  ungroup()

chisq_test <- chisq.test(table(adre_olink_MT_clustered$Cluster, adre_olink_MT_clustered$MAKE.90d))
p_value <- ifelse(chisq_test$p.value < 0.001, "< 0.001", format.pval(chisq_test$p.value, eps = 1e-3))

donut_data <- donut_data %>%
  mutate(MAKE.90d = factor(MAKE.90d, levels = c("N", "Y"))) %>%
  arrange(Cluster, MAKE.90d) %>%
  mutate(Label = paste0(MAKE.90d, "=", n, " (", round(Percentage, 1), "%)"))

cluster_colors <- my_palette
cluster_shades <- c(
  scales::alpha(cluster_colors[1], c(0.6, 1)),
  scales::alpha(cluster_colors[2], c(0.6, 1))
)

outer_ring <- donut_data %>% filter(Cluster == "1")
inner_ring <- donut_data %>% filter(Cluster == "2")

donut_plot <- ggplot() +
  geom_arc_bar(data = outer_ring, aes(x0 = 0, y0 = 0, r0 = 0.8, r = 1.2, fill = Label, amount = Percentage), stat = "pie", color = NA) +
  scale_fill_manual(values = cluster_shades[1:2], name = "MAKE90") +
  new_scale_fill() +
  geom_arc_bar(data = inner_ring, aes(x0 = 0, y0 = 0, r0 = 0.4, r = 0.8, fill = Label, amount = Percentage), stat = "pie", color = NA) +
  scale_fill_manual(values = cluster_shades[3:4], name = "MAKE90") +
  coord_fixed() +
  theme_void() +
  theme(
      legend.title = element_text(size = 17, face = "bold"),
      legend.text = element_text(size = 17),
      legend.position = "bottom",
      legend.box = "vertical"
  ) +
  annotate("text", x = 0, y = 0, label = paste0("p ", p_value), size = 6, color = "black")

donut_plot
```

### 4.1.4) Combining plots

```{r Combine Plots}
my_palette <- c("1"="#002395", "2"="#EF4135")

cohort_name <- "ADRENOSS Cohort"
n1 <- sum(adre_cox_df$Cluster == "1")
n2 <- sum(adre_cox_df$Cluster  == "2")

p_label <- ggdraw() +
  geom_rect(aes(xmin = 0.20, xmax = 0.80, ymin = 0.75, ymax = 0.95), fill = "lightgray", color = "black", size = 0.5) +
  draw_label("ADRENOSS Cohort", fontface = 'bold', size = 28, x = 0.5, y = 0.85, color = "red") +
  draw_label(paste0("Subphenotype 1 (n=", n1, ", ", round(n1 / (n1 + n2) * 100), "%)"), fontface = 'bold', size = 20, x = 0.5, y = 0.5, lineheight = 1.3, color = my_palette["1"]) +
  draw_line(x = c(0.1, 0.2), y = c(0.5, 0.5), color = my_palette["1"], linewidth = 2) +
  draw_label(paste0("Subphenotype 2 (n=", n2, ", ", round(n2 / (n1 + n2) * 100), "%)"), fontface = 'bold', size = 20, x = 0.5, y = 0.2, lineheight = 1.3, color = my_palette["2"]) +
  draw_line(x = c(0.1, 0.2), y = c(0.2, 0.2), color = my_palette["2"], linewidth = 2)

survival_panel <- cowplot::plot_grid(sur_plot$plot, sur_plot$table, ncol = 1, rel_heights = c(2, 1))

donut_plot_small <- wrap_elements(donut_plot) + plot_layout(widths = c(0.8))

cohort_panel <- (p_label | pca_plot) / (survival_panel | donut_plot_small) +
  plot_layout(heights = c(0.5, 1), widths = c(1.5, 0.7))


cohort_panel

# Save
save_plot(cohort_panel, "cohort.png", w = 14, h = 7)
```

# 5) KEGG / GO enrichment (ORA)

```{r kegg-go}
list_biomarker_from_olink <- readxl::read_xlsx("data/ProteinlistOlink-Dec14.xlsx") %>%
    mutate_if(.predicate = is.numeric, ~case_when(is.na(.) ~ 0, TRUE ~ .)) %>%  
    mutate_if(is.character, ~case_when(is.na(.) ~ "Missing", TRUE ~ .)) %>% 
    mutate(UniProt = case_when(UniProt == "NT-proBNP" ~ "NTproBNP", TRUE ~ UniProt)) %>%  
    distinct(UniProt, .keep_all = TRUE) %>% 
    dplyr::rename(ProteinName = "Protein name")

adre_olink_clustered_long <- read.csv("data/ADRENOSS_Olink_Long_deid.csv") %>%
  dplyr::filter(UniProt %in% biomarker_list_sig) %>%
  inner_join(adre_olink_MT_clustered %>% dplyr::select(SampleID, LCA.All), by = "SampleID") %>%  
  dplyr::rename(Cluster = LCA.All) 

# biomaRt with retry (internet-dependent)
repeat {
  ensembl <- tryCatch(useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl"), error = function(e) NULL)
  if (!is.null(ensembl)) break
  message("biomaRt connection failed, retrying in 5 seconds..."); Sys.sleep(5)
}

genes <- unique(adre_olink_clustered_long$Assay)
gene_ids <- getBM(attributes = c("hgnc_symbol","ensembl_gene_id","entrezgene_id"),
                  filters = "hgnc_symbol",
                  values = genes,
                  mart = ensembl) %>%
  distinct(hgnc_symbol, .keep_all = TRUE)

adre_olink_clustered_long <- adre_olink_clustered_long %>%
  dplyr::select(SampleID, Cluster, everything()) %>%
  mutate(Cluster = factor(Cluster)) %>% 
  inner_join(gene_ids, by = c("Assay" = "hgnc_symbol"))

wilcox_results <- olink_wilcox(
  df = adre_olink_clustered_long %>% dplyr::select(SampleID, Assay, UniProt, NPX, Cluster, OlinkID, Panel),
  variable = "Cluster", 
  alternative = "two.sided"
) %>% 
  dplyr::select(Assay, UniProt, OlinkID, Panel, p.value, Adjusted_pval, Threshold) %>%
  arrange(Adjusted_pval)

adr_pathway_data <- adre_olink_clustered_long %>%
  group_by(Cluster, UniProt, GeneName, entrezgene_id) %>%  
  summarize(Mean_NPX = mean(NPX), .groups = 'drop') %>%
  tidyr::pivot_wider(names_from = Cluster, values_from = Mean_NPX, names_prefix = "Cluster_") %>%
  mutate(Log2FC = Cluster_1 - Cluster_2) %>%  
  inner_join(wilcox_results %>% dplyr::select(UniProt, Adjusted_pval), by = "UniProt")

all_significant_genes_filtered <- adr_pathway_data %>% filter(Adjusted_pval < 0.05) %>% pull(entrezgene_id) %>% unique()
cl1_significant_genes_filtered <- adr_pathway_data %>% filter(Adjusted_pval < 0.05 & Log2FC > 0) %>% pull(entrezgene_id) %>% unique()
cl2_significant_genes_filtered <- adr_pathway_data %>% filter(Adjusted_pval < 0.05 & Log2FC < 0) %>% pull(entrezgene_id) %>% unique()

adre_KEGG <- enrichKEGG(
  gene = all_significant_genes_filtered, organism = "hsa",
  minGSSize = 10, pvalueCutoff = 0.05, pAdjustMethod = "BH"
) %>% setReadable(., org.Hs.eg.db, keyType = "ENTREZID")


# (Optional) GO examples
adre_GO  <- enrichGO(gene = cl1_significant_genes_filtered, OrgDb = org.Hs.eg.db,
                     keyType = "ENTREZID", ont = "BP", minGSSize = 10,
                     pvalueCutoff = 0.05, pAdjustMethod = "BH") %>%
            setReadable(., OrgDb = org.Hs.eg.db, keyType = "ENTREZID")

fc_vector <- adr_pathway_data$Log2FC; names(fc_vector) <- adr_pathway_data$GeneName
fc_vector <- fc_vector[!is.na(fc_vector) & !duplicated(names(fc_vector))]

p_kegg_dot <- enrichplot::dotplot(adre_KEGG, showCategory = 20) + ggtitle("KEGG Dotplot")
p_kegg_emap <- enrichplot::emapplot(pairwise_termsim(adre_KEGG), showCategory = 20, layout = "kk") + ggtitle("KEGG Emap")

kegg_plot <- (p_kegg_dot | p_kegg_emap) + patchwork::plot_layout(widths = c(.8, 1.2))

kegg_plot
# Save
save_plot(kegg_plot, "kegg_plot.png", w = 8, h = 10)
```

# 6) Volcano + PAC + Figure 1

## 6.1) Volcano Plot
```{r Volcano plot}
# EnhancedVolcano on combined table
deseq.volcano <- function(res, datasetName, print_plot = TRUE) {
  res$Adjusted_pval_log10 <- -log10(pmax(res$Adjusted_pval, 1e-10))
  p <- EnhancedVolcano(
    res, x = 'Log2FC', y = 'Adjusted_pval', lab = rownames(res),
    title = paste(datasetName, "KEGG Pathway Analysis"),
    subtitle = bquote(italic('FDR <= 0.05 and absolute FC >= 2')),
    labSize = 3, pointSize = 1.5, axisLabSize = 10,
    titleLabSize = 12, subtitleLabSize = 8, captionLabSize = 10,
    legendPosition = "none", pCutoff = 0.05, FCcutoff = 1
  )
  if (print_plot) print(p)
  invisible(p)
}

# Volcano using olink_volcano_plot
wilcox_results_vol_adre <- wilcox_results %>%
  inner_join(adr_pathway_data %>% dplyr::select(UniProt, Log2FC), by = "UniProt") %>%
  dplyr::rename(estimate = "Log2FC")

vol_plot <- wilcox_results_vol_adre %>%
  olink_volcano_plot() +
  geom_text_repel(aes(label = UniProt), max.overlaps = 0, fontface = "bold") +  
  labs(title = NULL, x = expression(Log[2]*"FC"), y = expression(-Log[10]*"(p-value)")) + 
  theme(
    plot.title = element_text(color = "red", face = "bold", size = 26), 
    legend.position = c(0.98, 0.02),
    legend.justification = c("right", "bottom"),
    legend.background = element_rect(fill = "white", color = "black"),
    legend.box.margin = ggplot2::margin(t = 0),
    legend.title = element_blank(),
    legend.text = element_text(size=21, color = "black"),
    axis.title = element_text(size = 22, face = "bold", color = "black"),
    axis.text = element_text(color = "black", size = 21)
  )

vol_plot 
```

## 6.2) PAC plot

```{r PAC plot}
k_vals <- 2:6
pac_df <- data.frame(k = k_vals, PAC = adre_pac*100)
best_row <- pac_df[which.min(pac_df$PAC), ]

p_pac <- ggplot(pac_df, aes(x = k, y = PAC)) +
  geom_line(color = "blue", linewidth = 1.1) +
  geom_point(color = "red", size = 3) +
  geom_point(data = best_row, aes(x = k, y = PAC), shape = 21, fill = NA, size = 4, stroke = 1, color = "blue") +
  geom_vline(xintercept = best_row$k, linetype = "dashed", color = "red", linewidth = .8) +
  labs(x = "Number of Clusters (k)", y = "PAC Score") +
  ylim(30, 65) +  
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    axis.line = element_line(color = "black"),
    axis.text = element_text(color = "black", size = 16),
    axis.title = element_text(color = "black", size = 20, face = "bold"),
    legend.position = "none",
    plot.title = element_blank()
  )
p_pac
```

## 6.3) Figure 1 — Combine volcano + heatmap + PAC

```{r Figure 1 - Combine Plot}
# Use the ggplotified heatmap
heatmap_plot_gp <- heatmap_plot_gp

theme_tag <- theme(plot.tag = element_text(size = 10, face = "bold", color = "black"))
p_pac2 <- p_pac + theme_tag + labs(tag = "A")
heatmap_plot2 <- heatmap_plot_gp + theme_tag + labs(tag = "B")
vol_plot2 <- vol_plot + theme_tag + labs(tag = "C")

combined_plot <- p_pac2 / heatmap_plot2 / vol_plot2 +
  plot_layout(heights = c(2.5, 4, 11))

combined_plot

# Save
save_plot(combined_plot, "combined_plot.png", w = 8, h = 10)
```

```{r session-info}
sessionInfo()
```
